# Goals, Power, Sample Size

```{r ch13-setup, include = FALSE}
require(mosaic)
require(brms)
require(CalvinBayes)
options(mc.cores = parallel::detectCores())
options(mc.cores = 1)
```

## Intro

### Goals

Kruschke (bold added):

> Any goal of research can be formally expressed in various ways. In this
chapter I will focus on the following goals formalized in terms of the highest
density interval (HDI):

> * Goal: **Reject a null** value of a parameter.
<br><br>Formal expression: Show that a region of practical equivalence (ROPE)
around the null value excludes the posterior 95% HDI.<br><br>
* Goal: **Affirm a predicted value** of a parameter.
<br><br> Formal expression: Show that a ROPE around the predicted value includes
the posterior 95% HDI.<br><br>
* Goal: **Achieve precision** in the estimate of a parameter.
<br><br>Formal expression: Show that the posterior 95% HDI has width less than a
specified maximum.<br><br>

### Obstacles

> The crucial obstacle to the goals of research is that 
**a random sample is only a probabilistic representation of the population**
from which it came.
Even if a coin is actually fair, a random sample of flips will rarely show 
exactly 50% heads. And even if a coin is not fair, it might come up heads 
5 times in 10 flips.
Drugs that actually work no better than a placebo might happen to cure
more patients in a particular random sample. And drugs that truly are effective
might happen to show little difference from a placebo in another particular
random sample of patients. 
Thus, 
**a random sample is a fickle indicator of the true state of the underlying world**.
Whether the goal is showing that a suspected value is or isn't credible, or
achieving a desired degree of precision, random variation is the researcher’s
bane.
**Noise is the nemesis**.

## Power

> Because of random noise, the goal of a study can be achieved only
probabilistically. The probability of achieving the goal, given the hypothetical
state of the world and the sampling plan, is called the **power** of the planned
research.

### Three ways to increase power

1. **Reduce measurement noise** as much as possible. 

    We can avoid sources of variability by controlling them (keep them
    constant for all observational units) or by including them as co-variates
    inour model.
    
    "Reduction of noise and control of other influences is the primary 
    reason for conducting experiments in the lab instead of in the 
    maelstrom of the real world." (But with the downside that lab 
    conditions might not accurately reflect the conditions of the situation
    we are actually interested in.)

2. **Amplify** the underlying **magnitude of the effect** if we possibly can. 

    Exmples: In a study of methods to improve reading performance, we might 
    select participants who are reading below grade level and so have 
    the most room to improve.  When administering a drug, a larger
    dose is likely to show a larger effect (within reason -- we don't 
    want undo side effects).

3. **Increase sample size**.

    In principle, more data means more information and less variability.
    In practice, more data means more costs (time, money, etc.).

## Calculating Power: 3 step process

We can calculate the power of our study by repeating the following three
steps many times:

1. **Sample parameters from a hypothetical distribution** of parameter values.

    These are chosen to reflect some situation we are interested in. Our main
    question will be "How likely are we to attain our goals in this situation."
    
2. Given the sampled parameter values, **simulate a data set** 
(using the data collection method proposed for the study).

3. Using the data set, **compute the posterior HDI** (or some other measure
related to our goal).

Each HDI can be classified as achieving or not achieving the goal.
**The proportion of simulated data sets that achieve the goal is our estimate of power,**
given the hypothesized distribution of parameter values.


## Power Examples

### Example: Catching an unfair coin

Let's go through these steps to estimate the power of detecting a 
coin that comes up heads 65% of the time using a ROPE of (0.49, 0.51).
Our goal in this case is an HDI that is either below 0.49 or above 0.51.

#### Sample parameters

In this case, this is pretty boring since our hypothetical situation is that
the proportion is exactly 0.65. But for the sake of what is to come, we will 
create a little function to compute 0.65 for us. To improve output formatting,
and so things generalize later, we will return that value as a named vector.
(A data frame or list would also work.)

We can test it out by "doing" it 3 times.

```{r ch13-draw-theta-coins}
library(mosaic)              # mosaic contains the do() function
draw_theta <- function() { c(theta = 0.65) }
do(3) * draw_theta()
```

#### Simulate data

For a given parameter value, we need to simulate data, in this case a number of heads
and a number of flips. In our example, the number of flips is always the same,
but we could design the study differently so that the number of flips was not always
the same, for example by flipping until we see a certain number of heads (or
tails, or both). We will make this an argument to our function so we can explore
different sample sizes later.

```{r ch13-simulate-data-coins}
simulate_data <- 
  function(theta, n) {
    x <- rbinom(1, n, theta)
    data.frame(theta = theta, heads = x, flips = n)  
  }
do(3) * simulate_data(0.60, 100)
```

#### Compute the HDI

From our simulated data, we need to compute and HDI. To avoid needing to simulate,
we will use what we know about beta distributions to compute the posterior distribution
analytically. To keep the code simple, we will use a central 95% interval that is not
an HDI, but it should be close.

The default prior is uniform (${\sf Beta}(1,1)$), but we will allow arguments to 
the function that let us experiment with different priors as well.

```{r ch13-compute-hdi-coins}
compute_hdi <- function(data, prior = list(a = 1, b = 1)) {
  # posterior
  post <- data.frame(a = prior$a + data$heads, b = prior$b + data$flips - data$heads)
  # not quite the HDI
  data.frame(
    theta = data$theta,
    x = data$heads,
    n = data$flips,
    lo = qbeta(0.025, post$a, post$b),
    hi = qbeta(0.975, post$a, post$b)
  )
}
draw_theta() %>%
  simulate_data(n = 100) %>% 
  compute_hdi()
```


#### Lather, rinse, repeat

Now we repeat these three steps many times.

```{r ch13-power-coins-100, cache = TRUE}
do(5) * {
  draw_theta() %>%
    simulate_data(n = 100) %>%
    compute_hdi()
}
Sims100 <- 
  do(5000) * {
    draw_theta() %>%
      simulate_data(n = 100) %>%
      compute_hdi()
  }
Sims100 <-
  Sims100 %>% mutate(reject = (hi < 0.49) | (lo > 0.51))  
df_stats(~reject, data = Sims100, props, counts)
```

The power increases if we incrase the sample size.

```{r ch13-power-coins-200, cache = TRUE}
Sims200 <- 
  do(5000) * {
    draw_theta() %>%
      simulate_data(n = 200) %>%
      compute_hdi() 
  }
Sims200 <-
  Sims200 %>%  mutate(reject = (hi < 0.49) | (lo > 0.51)) 

df_stats(~reject, data = Sims200, props, counts)
```

### Example: Estimating a Proportion

Suppose we would like to estimate a proportion within $\pm 2$%.
Further suppose we have no idea what that proportion is.
So now we will draw values of $\theta$ from a uniform distribution.


#### 

```{r ch13-draw-theta-precision}
draw_theta_unif <- function(a = 0, b = 1) { c(theta = runif(1, a, b)) }
do(3) * draw_theta_unif()
```
```{r ch13-power-precision-200, cache = TRUE}
Sims200p <- 
  do(5000) * {
    draw_theta_unif() %>%
      simulate_data(n = 200) %>%
      compute_hdi()
  }
Sims200p <-
  Sims200p %>% 
  mutate(
    width = hi - lo,
    success = width < 0.02) 

df_stats(~ success, data = Sims200p, props, counts)
```

So a sample of size 200 isn't going to do it (at least not very often).
Let's try 2000.

```{r ch13-power-precision-2000, cache = TRUE}
Sims2000p <- 
  do(5000) * {
    draw_theta_unif() %>%
      simulate_data(n = 2000) %>%
      compute_hdi()
  }
Sims2000p <-
  Sims2000p %>% 
  mutate(
    width = (hi - lo),
    center = lo + width / 2,
    success = width < 0.02) 

df_stats( ~ success, data = Sims2000p, props, counts)
```

Better, but we're still hitting our target only 10% of the time.

```{r ch13-power-precision-8k, cache = TRUE}
Sims8kp <- 
  do(5000) * {
    draw_theta_unif() %>%
      simulate_data(n = 8000) %>%
      compute_hdi()
  }
Sims8kp <-
  Sims8kp %>% 
  mutate(
    width = (hi - lo),
    center = lo + width / 2,
    success = width < 0.02) 

df_stats( ~ success, data = Sims8kp, props, counts)
```

We can find out a bit more about what is going on here by comparing the width
of the interval to the simulate value of $\theta$.


```{r ch13-Sims-2000-coin}
gf_point(width ~ theta, data = Sims8kp, shape = 21) %>%
  gf_hline(yintercept =  0.02, color = "red")
```


As we see, the width of the interval is wider when $\theta$ is near 0.5.
This means we can get by with a smaller sample size if we have good reason
to expect that $\theta$ is near 0 or 1.  And we will need a larger 
sample size if we expect $\theta$ is near 0.5.


```{r ch13-power-precision-varying-sample-size, cache = TRUE}
Sims_small_theta <- 
  do(5000) * {
    draw_theta_unif(0,0.2) %>%
      simulate_data(n = 3000) %>%
      compute_hdi()
  }

Sims_small_theta <-
  Sims_small_theta %>% 
  mutate(
    width = hi - lo,
    success = width < 0.02
  ) 

df_stats(~success, data = Sims_small_theta, props, counts)
gf_histogram(~ width, data = Sims_small_theta, bins = 40)
gf_point(width ~ theta, data = Sims_small_theta, shape = 21)
```

### More Complex Example

The one proportion example is as simple as it gets


* only one parameter (the proportion)
* simple distribution (Bernoulli)
* relatively easy to come up with hypothetical distributions
* easy calculation of the posterior distribution

In a typical situation, none of these special aspects will hold. Let's try something just one step more complex: a simple linear model.

```{r ch13-slr-draw-theta}
draw_theta_slr <- function() {
  data.frame(intercept = runif(1, 10, 12), 
             slope = runif(1, 3, 4),
             sigma = runif(1, 4, 5)
  )
}

do(3) * draw_theta_slr()
```


```{r ch13-slr-simulate-data}
simulate_data_slr <- 
  function(theta = list(slope = 1, intercept = 0, sigma = 1), n = 40, xlim = c(0,1)) {
    tibble(
      x = runif(n, xlim[1], xlim[2]),
      y = theta$intercept + theta$slope * x + rnorm(n, 0, theta$sigma)
    )
  }

gf_point(y ~ x, data = simulate_data_slr())
gf_point(y ~ x, 
         data = simulate_data_slr(
           xlim = c(0,10), 
           theta = list(slope = 2, intercept = 10, sigma = 0.5))
)
gf_point(y ~ x, 
         data = simulate_data_slr(
           xlim = c(0,10), 
           theta = list(slope = 2, intercept = 10, sigma = 2.5))
)
```

```{r ch13-slr-brm, cache = TRUE, results = "hide"}
slr_model <- brm( y ~ x, data = simulate_data_slr())
```

```{r ch13-slr-hdi}
compute_hdi_slr <-
  function(data) {
    model <- update(slr_model, newdata = data)
    model %>% posterior() %>% hdi()
  }
```

```{r ch13-slr-once, results = "hide"}
draw_theta_slr() %>%
  simulate_data_slr() %>%
  compute_hdi_slr()
```

Let's do 100 simulations.  This is going to require posterior sampling from 100
models, so we want to be confident things are working properly before we simulate
thousands of these.  The use of `update()` will make this run *much* faster.

```{r ch13-slr-do, results = "hide", cache = TRUE}
Sims_slr <- 
  do(100) * {
    draw_theta_slr() %>%
      simulate_data_slr() %>%
      compute_hdi_slr()
  }
```

Here is a plot showing the simulated HDIs. For any particular goal, it would 
now be easy to compute how often the goal is attained.

```{r ch13-slr-plot}
gf_pointrange( mode + lo + hi ~ .index, data = Sims_slr) %>%
  gf_facet_wrap( ~ par, scales = "free")
```
