
# Estimating One and Two Means

## Basic Model for Two Means

### Data

Two variables

  * metric reponse
  * dichotomous explanatory
 
### Model

The traditional starting point for modeling means is to assume that 
the each group is sampled from a normal distribution 
with unknown mean and a **common standard deviation**.
(We'll see that is no harder to have different standard deviations.)

So for two groups our model has three parameters: two means ($\mu_1$
and $\mu_2$) and one standard deviation $\sigma$.  

Of course, we also need priors for these parameters.
A common prior for the means is a normal distribution.
We will start with a uniform prior for the standard deviation,
but discuss better alternatives shortly.

That gives the following template for our model:

\begin{align*}
Y_{ij} & \sim {\sf Norm}(\mu_i, \sigma)
\\
\mu_i & \sim {\sf Norm}(?, ?)
\\
\sigma & \sim {\sf Unif}(?, ?)
\end{align*}

The questions marks will be filled in based on considerations of 
the scale (order of magnitude of the data) and the amount of regularizing
we want to do.

## An Old Sleep Study

Cushny, A. R. and Peebles, A. R. (1905) 
"The action of optical isomers: II hyoscines." 
*The Journal of Physiology* 32, 501–510.

**Design**: Subjects sleep habits were compared without a sleep inducing
drug and then with to see how two different drugs affected sleep.

### Data

Let's look at the data.  (`extra` = additional sleep on drug; `group` should 
really be `drug`, so let's rename it.)

```{r}
library(ggformula)
library(dplyr)
sleep <- 
  datasets::sleep %>% rename(drug = group)
gf_boxplot(extra ~ drug, data = sleep)
 df_stats(extra ~ drug, data = sleep)
```


### Model

It is simple enough to convert the model description above into a JAGS model,
but we need to fill in those question marks.  Let's try this:

* mean for prior on $\mu_i$: 0
    * corresponds to the drug having no impact on sleep
    * allows drug to increase or decrease sleep without prejudice
    * any other number would require more justification
    * will tend to pull estimates toward 0 (shrinkage) -- we are requiring
    evidence to convince us that the drug does something to sleep.

* sd for prior on $\mu_i$: 3

    * Says we are 95% certain that the average impact of a drug will be betweeen
    -6 and 6 additional hours of sleep and that it is very unlikely the drug
    will change sleep by 9 or more hours.  This is fairly week prior (6 extra
    hours of sleep would be a lot).  This might be chosen in consultation
    with scientists who are more familiar with what is reasonable.
    
* range for $\sigma$: One crude way to set the prior is to give a ball mark
estimate for the the standard deviation of the amount of sleep change in each
treatment group and then make sure we cover a range 3 orders of maginitude in
each direction.

* We can experiment with different priors to see how the impact results.

```{r, results = "hide"}
library(R2jags)
sleep_model <- function() {
  for (i in 1:Nobs) {
    extra[i] ~ dnorm(mu[drug[i]], 1/sigma^2)
  }
  for (d in 1:Ndrugs) {
    mu[d] ~ dnorm(0, 1/3^2)          # sd = 3
  }
  sigma ~ dunif(2/1000, 2 * 1000)    # 3 orders of mag each way of 2
  delta <- mu[2] - mu[1]
  tau   <- 1 / sigma^2               
}

sleep_jags <- 
  jags(
    model = sleep_model,
    parameters.to.save = c("mu", "sigma", "delta", "tau"),
    data = list(
      extra = sleep$extra,
      drug  = sleep$drug,
      Nobs  = nrow(sleep),
      Ndrugs = 2
    ),
    DIC = FALSE  # because we haven't discussed deviance yet
  )
```

```{r}
sleep_jags
library(bayesplot)
sleep_mcmc <- as.mcmc(sleep_jags)
mcmc_areas(sleep_mcmc, prob = 0.95)
mcmc_pairs(sleep_mcmc)
mosaic::prop( ~(delta > 0), data = posterior(sleep_jags))
```

### Separate standard deviations for each group

```{r, results = "hide"}
sleep_model2 <- function() {
  for (i in 1:Nobs) {
    extra[i] ~ dnorm(mu[drug[i]], 1/sigma[drug[i]]^2)
  }
  for (d in 1:Ndrugs) {
    mu[d]    ~  dnorm(0, 1/3^2)
    sigma[d] ~  dunif(2/1000, 2 * 1000)  
    tau[d]   <- 1 / sigma[d]^2
  }
  delta <- mu[2] - mu[1]
}

sleep_jags2 <- 
  jags(
    model = sleep_model2,
    parameters.to.save = c("mu", "sigma", "delta", "tau"),
    data = list(
      extra = sleep$extra,
      drug  = sleep$drug,
      Nobs  = nrow(sleep),
      Ndrugs = 2
    ),
    DIC = FALSE
  )
```

```{r}
library(bayesplot)
library(CalvinBayes)
sleep_jags2
sleep_mcmc2 <- as.mcmc(sleep_jags2)
mcmc_areas(sleep_mcmc2, prob = 0.95)
mcmc_pairs(sleep_mcmc2)
mosaic::prop( ~(delta > 0), data = posterior(sleep_jags2))
hdi(sleep_jags2, pars = c("delta"))
hdi(sleep_jags2)
```

For those who know about 2-sample t tests:

```{r}
t.test(extra ~ drug, data = sleep)
```

### Paired Comparisons

The data actually contain another variable: `ID`.  As it turns out, the same
ten people were tested with each drug.  If we are primarily interested in 
comparing the two drugs, we might take the difference between the extra sleep 
with one drug and with the other drug **for each person**.  This is referred
to as a paired design.

A paired comparison of means is really just looking at one mean -- the mean difference.  We can do this a couple of different ways:

1. Compute the difference before giving data to JAGS
2. Build the differences into the JAGS code.

We will use option 1 here and convert our data so that each row corresponds
to one person and there are separate columns for the extra sleep produced
by each drug.  This is sometimes referred to as converting from **long format**
(more rows, fewer columns) to **wide format** (fewer rows, more columns).
The `tidyr::spread()` funciton is useful for this. (And `tidyr::gather()`
can be used to convert in the opposite direction.)

```{r}
library(tidyr)
sleep_wide <-
  datasets::sleep %>% 
  rename(drug = group) %>%
  mutate(drug = paste0("drug", drug)) %>%
  spread(key = drug, value = extra) 
sleep_wide
sleep_wide <-
  sleep_wide %>%
  mutate(delta = drug2 - drug1)
sleep_wide
gf_boxplot(~ delta, data = sleep_wide)
```

```{r}
sleep_model3 <- function() {
  for (i in 1:Nsubj) {
    delta[i] ~ dnorm(mu, 1 / sigma^2)
  }
  mu ~ dnorm(0, 2)
  sigma ~ dunif(2/1000, 2 * 1000)
  tau <- 1 / sigma^2
}

sleep_jags3 <- 
  jags(
    model = sleep_model3,
    parameters.to.save = c("mu", "sigma", "tau"),
    data = list(
      delta = sleep_wide$delta,
      Nsubj = nrow(sleep_wide)
    ),
    DIC = FALSE
  )
```

```{r}
library(bayesplot)
library(CalvinBayes)
sleep_jags3
sleep_mcmc3 <- as.mcmc(sleep_jags3)
mcmc_areas(sleep_mcmc3, prob = 0.95)
mcmc_pairs(sleep_mcmc3)
mosaic::prop( ~(mu > 0), data = posterior(sleep_jags3))
hdi(sleep_jags3, pars = c("mu"))
hdi(sleep_jags3)
```

## Looking at Likelihood

```{r}
likelihood <- function(mu, sigma, x, log = FALSE) {
  D <- tibble(
    x = x,
    l = dnorm(x, mu, sigma, log = log),
  )
  L <- prod(dnorm(x, mu, sigma))
  logL <- sum(dnorm(x, mu, sigma, log = TRUE))
  
  T <- tibble(x = mean(x), logL = logL, height = 1.2 * dnorm(0, 0, 1, log = log))

  cat(paste0("log likelihood: ", format(logL)), 
      "; mu: ", format(mu), 
      "; sigma: ", format(sigma), "\n")
  
  gf_segment(0 + l ~ x + x, data = D, color = "red") %>%
    gf_point(0 ~ x, data = D, color = "red") %>%
    gf_function(function(x) dnorm(x, mu, sigma, log = log)) 
}
```

```{r, eval = FALSE}
library(manipulate)
manipulate(
  likelihood(MU, SIGMA, c(8, 12), log = LOG) %>% 
    gf_lims(x = c(0, 20), y = c(NA, 0.5)),
  MU = slider(5, 15, 10, step = 0.2),
  SIGMA = slider(1, 10, 2, step = 0.2),
  LOG = checkbox(FALSE, "log likelihood")
)
```



## Exercises

1. Fit a model that compares the mean height for men and women
using the 30-year-olds in the `NHANES` data set (in the NHANES package)
and answer the following questions about your model.

    a. Explain your choice of priors.
    a. Does this sample provide enough evidence to conclude that 
    men are taller (on average)?
    a. Does this sample provide enough evidence to conclude
    that the standard deviation of height differs between
    men and women?
    
    ```{r}
    Thirty <- NHANES::NHANES %>% filter(Age == 30)
    df_stats(Pulse ~ Gender, data = Thirty)
    gf_dens( ~ Pulse, color = ~ Gender, data = Thirty, binwidth = 1)
    ```