---
output:
  pdf_document: default
  html_document: default
---

# Multiple Nominal Predictors

## Crop Yield by Till Method and Fertilizer

The data in `CalvinBayes::SplitPlotAgri` are from an agricultural study in which 
different tilling methods and different fertilizers were used and the crop yield 
was subsequently measured.

```{r ch20-splitplot-plot}
gf_point(Yield ~ Fert | ~ Till, data = SplitPlotAgri, alpha = 0.4, size = 4)
```

Here are two models. See if you can figure out what they are.
(How can you use R to check if you are correct?)

* What parameters does each model have?
* Write a formula that describes the model. Be sure to clarify what the variables
mean.
* How would you use each model to estimate the mean yield when using ridge tilling
and deep fertilizer?  (Imagine that you already have the posterior distribution in 
hand.)


```{r ch20-fert1, results = "hide", cache = TRUE, eval = FALSE}
fert1_brm <-
  brm(Yield ~ Till + Fert, data = SplitPlotAgri)
```

\vfill

```{r ch20-fert2, results = "hide", cache = TRUE, eval = FALSE}
fert2_brm <-
  brm(Yield ~ Till * Fert, data = SplitPlotAgri)
```

\vfill


\newpage

## Split Plot Design

There is a bit more to our story.  The study used 33 different fields.  Each field
was divided into 3 sections and a different fertilizer was applied to each of the 
three sections. (Which fertilizer was used on which section was determined at random.)
This is called a "split-plot design" (even if it is applied to things that are not
fields of crops).

It would have been possible to divide each field into 9 sub-plots and use all
combinations of tilling and fertilizer, but that's not how this study was done.
The tilling method was the same for the entire field -- likely because it was 
much more efficient to plow the fields this way. 

The plot below indicates that different fields appear to have different baseline
yields since the dots associated with one field tend to be near the top or bottom
of each of the fertilizer clusters.  We can add an additional variable to our
model to handle this situation.

```{r ch20-splitplot-plot-with-fields}
gf_point(Yield ~ Fert | ~ Till, data = SplitPlotAgri, alpha = 0.4, size = 4) %>%
  gf_line(group = ~Field)
```

```{r ch20-fert3, results = "hide", cache = TRUE}
fert3_brm <-
  brm(Yield ~ Till * Fert + Field, data = SplitPlotAgri)
```

```{r ch20-fert4, results = "hide", cache = TRUE}
fert4_brm <-
  brm(Yield ~ Till + Fert + Field, data = SplitPlotAgri)
```
