
# Count Response

```{r ch24-setup, include = FALSE}
library(CalvinBayes)
library(brms)
library(loo)
```

## Hair and eye color data

```{r ch24-table}
HairEyeColor %>%
  tidyr::spread(Eye, Count)

gf_tile(Count ~ Hair + Eye, data = HairEyeColor) %>% 
  gf_refine(scale_fill_viridis_c(option = "C"))
```

## Are hair and eye color independent?

You probably suspect not.  We expect blue eyes to be more common among
blond-haired people than among black-haired people, perhaps. How do we 
fit a model to test if our intuition is correct using the data above?

If the rows and columns of the table were independent, then for each row $r$ and 
column $c$ the probability of being in a row $r$ and column $c$ would be the 
product of the probabilities of being in row $r$ and of being in column $c$:

$$
\begin{align*}
\frac{\mu_{r c}}{N} 
&=
\frac{y_{r\cdot}}{N} 
\cdot 
\frac{y_{\cdot c}}{N} 
\\
\mu_{r c}
&=
\frac{1}{N}
\cdot 
y_{r\cdot}
\cdot 
y_{\cdot c}
\\
\log(\mu_{r c})
&=
\underbrace{\log(\frac{1}{N})}_{\alpha_0}\cdot1
+  
\underbrace{\log(y_{r\cdot})}_{\alpha_{r \cdot}}\cdot1
+ 
\underbrace{\log(y_{\cdot c})}_{\alpha_{\cdot c}}\cdot1
\\
\log(\mu)
&=
\alpha_0
+  
\sum_{r = 1}^R \alpha_{r\cdot} [\![ \mathrm{in\ row\ } r ]\!]
+ 
\sum_{c = 1}^C \alpha_{\cdot c} [\![ \mathrm{in\ column\ } c ]\!]
\\
\log(\mu)
&=
\underbrace{(\alpha_0 + \alpha_{1\cdot} + \alpha_{\cdot 1})}_{\beta_0}
+  
\sum_{r = 2}^R \alpha_{r\cdot} [\![ \mathrm{in\ row\ } r ]\!]
+ 
\sum_{c=2}^C \alpha_{\cdot c} [\![ \mathrm{in\ column\ } c ]\!]
\\
\log(\mu)
&=
\beta_0
+  
\sum_{r = 2}^R \beta_{r\cdot} [\![ \mathrm{in\ row\ } r ]\!]
+ 
\sum_{c=2}^C \beta_{\cdot c} [\![ \mathrm{in\ column\ } c ]\!]
\end{align*}
$$

This looks exactly like our additive linear model (on the log scale) and so
the common name for this model is the **log linear model**.

If the rows and columns are not independent, then we will have non-zero interaction
terms indicating how far things are from independent. We know how to 
add in interaction terms, so we are good to go there.

All that remains is to come up with a good distribution that turns a mean $\mu$
into a count.  We don't expect the cell count $y_{rc} to be exactly $\mu_{rc}$
(especially when $\mu_{rc}$ in not an integer!). 
But values close to $\mu_{rc}$ should be more likely than values
farther away from $\mu_{rc}$.
A Poisson distribution has exactly these properties and makes a good model
for the noise in this situation.

Poisson distributions
have one parameter (often denoted $\lambda$) satisfying

$$
\begin{align*}
\mathrm{mean} &= \lambda \\
\mathrm{variance} &= \lambda \\
\mathrm{standard\ deviation} &= \sqrt{\lambda} \\
\end{align*}
$$

Here are several examples of Poisson distributions.  Notice that $\lambda$ 
need not be an integer, but all of the values produced by a Poisson random 
process are integers.

```{r ch24-poisson-dists}
gf_dist("pois", lambda = 1.8)
gf_dist("pois", lambda = 5.8)
gf_dist("pois", lambda = 25.8)
gf_dist("pois", lambda = 254.8)
```

The Poisson distributions become more and more symmetric as $\lambda$ increases.
In fact, they become very nearly a normal distribution.[^24-1]

[^24-1]: Take Stat 343 to find out why.

## Poisson model

The discussion above gives us enough information to create the appropriate 
model in R using `brm()`.

```{r ch24-poison-model, results = "hide", cache = TRUE}
color_brm <- 
  brm(Count ~ Hair * Eye, data = HairEyeColor, family = poisson(link = log))
```

```{r ch24-poison-model-look}
color_brm  
```

Our main question is whether any of the interaction terms are credibly different
from 0. That would indicate a cell that has more or fewer observations than
we would expect if rows and columns were independent. We can construct contrasts
to look at particular ways in which independence might fail.

```{r ch24-hair-eye-no-interaction, cache = TRUE, results = "hide"}
color2_brm <- 
  brm(Count ~ Hair + Eye, data = HairEyeColor, family = poisson(link = log))
```

The model with interaction has higher estimated elpd than the model
without interaction terms, an indication that there are credible interaction
effects.

```{r ch24-hair-eye-elpd}
loo_compare(waic(color_brm), waic(color2_brm))
```

(LOO gives a similar result, but requires starting from scratch for 
several observations, so it is slower.)

As an example, let's test whether blond-haired people are more likely to have blue
eyes than black-haired people. We don't want to compare counts, however, since 
the number of blond-haired and black-haired people is not equal. Differences
on a log scale are ratios on the natural scale. So we might compare 

$$
\begin{align*}
\log(\mu_{\mathrm{blond,\ blue}}) -  \log(\mu_{\mathrm{blond,\ not\ blue}})
&=
\log\left( \frac{\mu_{\mathrm{blond,\ blue}}}
           {\mu_{\mathrm{blond,\ not\ blue}}}\right)
\end{align*}
$$
with

$$
\begin{align*}
\log(\mu_{\mathrm{black,\ blue}}) -  \log(\mu_{\mathrm{black,\ not\ blue}})
&=
\log\left( \frac{\mu_{\mathrm{black,\ blue}}}
           {\mu_{\mathrm{black,\ not\ blue}}}\right)
\end{align*}
$$
If those two quantities are equal, then the log odds, hence odds, hence probability
of having blue eyes is the same in both groups.

Let's build the corresponding contrast and find out.  Since the intercept coefficient
shows up in every term (and then cancels out), we can drop it from our contrast to
save some typing. Similarly in the blond difference `b_HairBlond` drops
out and in the black-haired difference `b_HairBlack` drops out.
Things are further simplified because blue eyes and black hair are the reference
groups (because they come alphabetially first).

```{r ch24-color-interaction}
Post <- posterior(color_brm) 
names(Post)
Post <- Post %>%
  mutate(
    contrast = 
        0 - (b_EyeBrown + `b_HairBlond:EyeBrown` +
             b_EyeGreen + `b_HairBlond:EyeGreen` + 
             b_EyeHazel + `b_HairBlond:EyeHazel`) / 3 +  
      - 0 + (b_EyeBrown + b_EyeGreen + b_EyeHazel) / 3
  )
hdi(Post, pars = ~ contrast)
plot_post(Post$contrast)
```

As expected, the posterior distribution for this contrast is shifted well away from 0,
an indication that the proportion of blond-haired people with blue eyes is credibly
higher than the proportion of black-haired people with blue eyes.
The log odds ratio is about 2 (posterior HDI suggests somewhere betweeen 1.4 and 2.7).
and the odds ratio can be obtained by exponentiation.

```{r ch24-color-odds-ratio}
hdi(Post, pars = ~ exp(contrast))
plot_post(exp(Post$contrast))
```

Unfortunately, we can't convert the odds ratio directly into a relative risk.

$$
\begin{align*}
\mathrm{odds\ ratio} 
&= \frac{p_1 / (1-p_1)}{p_2 / (1-p_2)} \\
&= \frac{p_1}{1-p_1} \cdot \frac{1-p_2}{p_2} \\
&= \frac{p_1}{p_2}\cdot \frac{1-p_2}{1-p_1} \\
&= \mathrm{relative\ risk} \cdot \frac{1-p_2}{1-p_1} \\
\end{align*}
$$
Relative risk and odds ratio are numerically close when 
$\frac{1-p_2}{1-p_1}$ is close to 1, which happens when $p_1$ and $p_2$ are both
quite small.
