---
title: "Solutions to Problems in (Re)doing Bayesian Data Analysis"
author: "R Pruim"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: true
      smooth_scroll: true
---

<style>

body{
  color: darkslategray;
}

h1, h2, h3, h4 {
  color: black;
}

.solution{
  border: 2px solid navy;
  color: black;
  margin: 20px;
  padding: 8px;
  size: larger;
} 
</style>

```{r setup-sols, include = FALSE}
library(CalvinBayes)
library(ggformula)
library(mosaic)
library(R2jags)
library(bayesplot)
theme_set(theme_bw())
knitr::opts_chunk$set(
  cache = TRUE,
  fig.keep = "all",
  fig.show = "hold",
  fig.width = 9, fig.height = 3, out.width = "80%")
```

## Chapter 8


<!-- Exercise 8.4. [Purpose: Explore the prior on a difference of parameters implied from the priors on the individual parameters.] -->

1. You want to know who is the better free throw shooter, Alice or Bob.
You decide to have each shoot a number of shots and record their makes 
and misses. You are primarily interested in the difference between their
free throw shooting proportions ($\theta_2 - \theta_1$), and you are curious
to know how your choice of priors for $\theta_1$ and $\theta_2$ affects the
prior for $\theta_2 - \theta_1$.  For each situation below, use JAGS to 
sample from the prior distribution for $\theta_2 - \theta_1$ and create 
a density plot.  (In each case, assume the priors for $\theta_1$ and 
$\theta_2$ are independent.)

    a. Both priors are uniform.  What distribution do you think the 
    prior for $\theta_2 - \theta_1$ is?
    
    b. Both priors are ${\sf Beta}(0.2, 0.2)$.  Explain why the prior 
    for $\theta_2 - \theta_1$ looks the way it does.
    
<div class = "solution">

#### part a {-}

```{r sol08-priors-exercise-a}
# Let's do some renaming
Target <- z6N8z2N7 %>% rename(hit = y, subject = s)

# make a copy of our data list
TargetList0 <- list(
  Nobs = 0,  
  Nsub = 2,
  subject = as.numeric(as.factor(Target$subject))
)
bern2_model_a <- function() {
  for (i in 1:Nobs) {
    # each response is Bernoulli with the appropriate theta
    hit[i] ~ dbern(theta[subject[i]])  
  }
  for (s in 1:Nsub) {
    theta[s] ~ dbeta(1, 1)    # prior for each theta
  }
}
bern2_jags_a <-
  jags(
    data = TargetList0,
    model.file = bern2_model_a,
    parameters.to.save = c("theta"),
    DIC = FALSE)
bern2_jags_a
gf_dens( ~ (theta.2 - theta.1), data = posterior(bern2_jags_a))
```

Looks like a triangle distribution (and it is).

#### part b {-}

```{r sol08-priors-exercise-b}
bern2_model_b <- function() {
  for (i in 1:Nobs) {
    # each response is Bernoulli with the appropriate theta
    hit[i] ~ dbern(theta[subject[i]])  
  }
  for (s in 1:Nsub) {
    theta[s] ~ dbeta(.2, .2)    # prior for each theta
  }
}
bern2_jags_b <-
  jags(
    data = TargetList0,
    model.file = bern2_model_b,
    parameters.to.save = c("theta"),
    DIC = FALSE)
bern2_jags_b
gf_dens( ~ (theta.2 - theta.1), data = posterior(bern2_jags_b))
```

The priors for each $\theta$ say they are both likely to be close to 0 or 1.  If both are close to the same end, the difference will be small.  That's the central
peak.  If one is near 0 and the other near 1, then the difference will be near 
1 or -1, those are the two smaller peaks.  The entral peak is larger because there 
are two ways to land there 
($\theta_1 \approx 0$, $\theta_2 \approx 1$ or 
$\theta_2 \approx 0$, $\theta_1 \approx 1$).


</div>



2. Now suppose that Alice makes 25 out of 30 shots and Bob makes 18 out of 32.
Is this enough evidence to conclude that Alice is the better shooter?
Do this two ways.
In each case, use a {\sf Beta(4, 2)} prior.  What does this prior say about
what you believed about Alice and Bob's shooting before gather data?

    a. Create data and use a model like the one used elsewhere in this chapter.
    b. Instead of using `dbern()` use `dbin()` (JAGS version of the binomial distribution).  This should allow you to get by with simpler data that consists
    only of the numbers 25, 30, 18, and 32.

<div class = "solution">
#### part b {-}

```{r sol08-bball}
bball_model <- function() {
  for (s in 1:Nsub) {
    # each response is Bernoulli with the appropriate theta
    made[s] ~ dbin(theta[s], attempts[s])  
  }
  for (s in 1:Nsub) {
    theta[s] ~ dbeta(4, 2)    # prior for each theta
  }
}

bball_jags <- 
  jags(
    data = list(made = c(25, 18), attempts = c(30, 32), Nsub = 2),
    model.file = bball_model,
    parameters.to.save = c("theta")
  )
bball_post <-
  posterior(bball_jags) %>% 
  mutate(difference = theta.2 - theta.1)
hdi(bball_post, pars = "difference")
gf_dens( ~ difference, data = bball_post)
```
</div>

3. Let's think about Alice and Bob some more.
We don't know how to do this, but explain why if you knew very little about
basketball, you might like to have a prior for $\theta_1$ and $\theta_2$ 
that was not indpendent. 
How might the priors for $\theta_1$ and $\theta_2$ be related?

5. Consider the model below.

```{r, results = "hide"}
diff_model <- function() {
  z1 ~ dbin(theta1, n1)
  z2 ~ dbin(theta2, n2)
  theta2 <- theta1 + diff
  theta1 ~ dbeta(4, 2)
  diff ~ dnorm(0, 400)  # Normal with mean 0 and sd = 0.05
}

diff_jags <-
  jags.parallel(
    model = diff_model,
    data  = list(n1 = 30, z1 = 25, n2 = 32, z2 = 18),
    parameters.to.save = c("theta1", "theta2", "diff"),
    n.iter = 5000,
    n.chains = 4
  )
```

    a. Create a plot of the **prior** distribution of 
    `theat1` and `theta2`.  A scatter plot with overlayed density plot
    works well. How does this prior compare to the priors in problem
    1. (Hint: Don't forget to use `DIC = FALSE`.)
    
    b. Now fit the model to the Alice and Bob data.  How does this
    choice of prior change things?
    
```{r sol08-diff-model}
diff_jags0 <-
  jags.parallel(
    model = diff_model,
    data  = list(n1 = 0, z1 = 0, n2 = 0, z2 = 0),
    parameters.to.save = c("theta1", "theta2", "diff"),
    DIC = FALSE
  )

diff_jags0
gf_dhistogram(~diff, data = posterior(diff_jags0))
gf_point(theta2 ~ theta1, data = posterior(diff_jags0), alpha = 0.3) %>%
  gf_density2d()
gf_point(theta2 ~ theta1, data = posterior(diff_jags), alpha = 0.3) %>%
  gf_density2d()
mcmc_areas(as.mcmc(diff_jags0))
mcmc_areas(as.mcmc(diff_jags), pars = c("theta1", "theta2", "diff"))
```